---
title: "20230905_twoLine_model.R"
output:
  html_document:
    df_print: paged
date: "2023-09-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Two-line model

Assume a model where the percentage of disease changes according to the addition of two lines. The first has a rate \>= 0, the second with a rate \<= 0 starts at some time after the initial time point of the first line.

(Currently, both lines use a first-order rate; an alternative model would be that the second effect depends on level at the given time point)

```{r cars}
rm(list=ls())
cat('\014')
graphics.off()

suppressWarnings({      #no warnings to avoid cluttering Knit
library('tidyverse',quietly=TRUE,warn.conflicts = FALSE,verbose=FALSE) 
library('rjags')
source('./DBDA2E-utilities_mod.R')  #modified to fit markdown
source('./20230912_twoLine_inclFac_fit.R')
})
```

## Model visualization and SSQ parameter estimate

```{R}
set.seed(01092023)

#testCase= c(a0=2, a1=1, a2= -1.5, t0=40, noiseSD= 0.1)
#testCase= c(a0=2, a1=1, a2= -0.05,t0=40, noiseSD= 0.1)
#testCase= c(a0=2, a1=-1, a2= 0.5,t0=40, noiseSD= 0.1)
#testCase= c(a0=2, a1=-0.1, a2= 0.5,t0=40, noiseSD= 0.1)
testCase= c(a0=2, a1=1, a2= -0.5,t0=40, noiseSD= 0.3)
a0= testCase[1]; a1=testCase[2]; a2= testCase[3]
t0= testCase[4]
noiseSD= testCase[5]

t= 0:100

y_noNoise= a0 + a1*t + a2*ifelse(t>t0,t-t0,0)
#noise1= rnorm(length(t),mean=0,sd=1)  #constant noise level
noise2= rnorm(length(t),mean=0,sd=noiseSD*abs(y_noNoise))  #noise relative to data
y= y_noNoise+noise2
y[y<0]= 0   #make sure no negative values
DF= cbind.data.frame(t=t,y=y)

plot(t,y,ylim=c(0,1.1*max(y)))

twoLine_fun= function(t, pms, obs){  #ssq minimization between observed data
                                       #and two-line model
  a0= pms[1]; a1=pms[2]; a2=pms[3]; t0= pms[4]
  y= a0 + a1*t + a2*ifelse(t>t0,t-t0,0)
  
  ssq= sum( (obs-y)^2 )
  return(ssq)
}

pms= c(a0,a1,a2,t0)
#twoLine_fun(t,pms,obs=y)

nlmOUT= nlm(f=twoLine_fun,p=c(a0=0,a1=1,a2=1,t0=max(t)/2),t=t,obs=y)
est= nlmOUT$estimate

lines(t, est[1] + est[2]*t + est[3]*ifelse(t>est[4],t-est[4],0),col='blue')
```

# Estimate parameters based on Bayesian model
```{R}
DF1= twoLine_fit(t=t,y=y)
head(DF1)
```

```{R}
# MCMC results
par(mfrow=c(2,2))
hist(DF1$b0,col='skyblue',breaks=50)
lines(x=HDIofMCMC(DF1$b0,credMass=.95),y=c(0,0),lwd=5)

hist(DF1$b1,col='skyblue',breaks=50)
lines(x=HDIofMCMC(DF1$b1,credMass=.95),y=c(0,0),lwd=5)

hist(DF1$b2,col='skyblue',breaks=50)
lines(x=HDIofMCMC(DF1$b2,credMass=.95),y=c(0,0),lwd=5)

hist(DF1$t0,col='skyblue',breaks=50)
lines(x=HDIofMCMC(DF1$t0,credMass=.95),y=c(0,0),lwd=5)


par(mfrow=c(1,1))
plot(DF1$b1,DF1$b2,col='skyblue')

par(mfrow=c(1,2))
hist(DF1$iF1,xlim=c(-.2,1.2),breaks=seq(-0.1,1.1,.2),col='skyblue',xaxt='n')
axis(side=1, at=c(0,1), labels=c(0,1))
hist(DF1$iF2,xlim=c(-.2,1.2),breaks=seq(-0.1,1.1,.2),col='skyblue',xaxt='n')
axis(side=1, at=c(0,1), labels=c(0,1))
```

```{R}
# Posterior prediction
n= 20
y_pred= matrix(NA,nrow=n,ncol=length(t))
pred_idx= sample(1:dim(DF1)[1],size=n)  #posterior prediction for n values
for(i in 1:length(pred_idx)){
  #print(paste(i))
  y_pred[i,]= DF1$b0[i] + DF1$iF1[i]*DF1$b1[i]*t + ifelse(DF1$t0[i]>=t,DF1$iF2[i]*DF1$b2[i]*(t-DF1$t0[i]),0)
}
plot(type='n',x=t,y=rep(0,length(t)),ylim=c(0,max(y_pred)))
for(i in 1:dim(y_pred)[1]){
  lines(t,y_pred[i,])
}
```

# Test prediction power for some simulated data
 (no systematic evaluation)
## Rate constants different from 0 (evaluate sensitivity)
```{R}
a0=2
t= 0:30 #match the actual data: 1 measurement per year for 40 years
t0= 20
coefDF= expand.grid(a1=c(0.01,0.1,1), a2=c(-.01,-.1,-0.5))   #a1>0, a2<0


#dev.new(width=12,height=12,noRStudioGD = TRUE)
par(mfrow=c(nrow=ceiling(dim(coefDF)[1]/3),ncol=3),mar=c(3,3,1,0))
DF_result= data.frame(matrix(NA,nrow=dim(coefDF)[1],ncol=8))
colnames(DF_result)= c('b1','b2','b1_correct','b2_correct','b1_relErr', 'b2_relErr','est_b1','est_b2')
for(i in 1:dim(coefDF)[1]){  
  a1= coefDF$a1[i]; a2= coefDF$a2[i]
  DF_result[i,c(1,2)]= c(a1,a2)
  y_noNoise= a0 + a1*t + a2*ifelse(t>t0,t-t0,0)
  noise= rnorm(length(t),mean=0,sd=0.1*abs(y_noNoise))  #noise relative to data
  y= y_noNoise+noise
  y[y<0]= 0   #make sure no negative values
  plot(t,y,ylim=c(0,1.1*max(y)),main=paste('b1:',a1,', b2:',a2)) #change naming: 'a' are simulated, 'b' estimated params
  
  DF1= twoLine_fit(t,y)
  
  b1_correct= ifelse( mean(DF1$iF1)>.8, 1, 0)  #indicator for b1==1 at least 80% of time
  b1_relErr= abs(mean(DF1$b1)-a1)/abs(a1+1E-3)  #relative error of mean estimate; w. flooring
  b2_correct= ifelse( mean(DF1$iF2)>.8, 1, 0)  #indicator for b2==1 at least 80% of time
  b2_relErr= abs(mean(DF1$b2)-a2)/abs(a2+1E-3)  #relative error of mean estimate; w. flooring
  DF_result[i,3:8]= c(b1_correct,b2_correct,b1_relErr,b2_relErr,median(DF1$b1),median(DF1$b2))

  lines(t, mean(DF1$b0) + mean(DF1$b1)*t + mean(DF1$b2) *ifelse(t>mean(DF1$t0),t-mean(DF1$t0),0), col='blue')
}

#manually reformat
DF_result_long= cbind.data.frame(
  coef= c(DF_result$b1,DF_result$b2),
  estCoef= c(DF_result$est_b1,DF_result$est_b2),
  x= factor(c(rep(1,dim(DF_result)[1]),rep(2,dim(DF_result)[1])),levels=c(1,2)),
  y= rep(1:dim(DF_result)[1],times=2),
  correct= factor(c(DF_result$b1_correct,DF_result$b2_correct),levels=c(0,1)),
  relErr= c(DF_result$b1_relErr,DF_result$b2_relErr),
  log10_recipRelErr= log(1/c(DF_result$b1_relErr,DF_result$b2_relErr),10) #log10 of recip. relErr
  )

ggplot()+
  geom_point(data=DF_result_long,aes(x=x,y=y,color=correct,size=relErr))+  #log10_recipRelErr
  scale_color_manual(limits=as.character(c(0,1)),values=c('red','green'),labels=c('no','yes'))+
  scale_size_continuous(trans='log10')+
    #breaks=log(1/c(0.01,0.1,1),10),labels=c(0.01,0.1,1),name='rel. Error',
     #                   limits=log(1/c(0.005,1.005),10))+
  geom_text(data=DF_result_long,aes(x=as.numeric(x)-0.2,y=y, label=coef))+
  theme_minimal()+
  scale_x_discrete(breaks=c(1,2),labels=c('b1','b2'))+
  theme(axis.title= element_blank(),axis.text.y=element_blank())
```

At a noise level of SD = 10%, a rate or 0.01 / -0.01 percent/year is generally
not detectable. Decline rates of -0.1 were also missed in 2 out of 3 cases and once even decline rate of -0.5 was not detected. The relative error is generally large only for small coefficients, but there are exceptions.
Note that start of decline was fixed to 20th of the 30 years; changing this
choice and changing assumed variation will alter estimation power.

## At least one rate constant == 0 (evaluate selectivity)
```{R}
a0=2
t= 0:30 #match the actual data: 1 measurement per year for 40 years
t0= 20
coefDF= data.frame(
  a1= c(0, 0.,  0.01,0.1),
  a2= c(0,-0.01,0,   0)
)   #a1>=0, a2<=0


#dev.new(width=12,height=12,noRStudioGD = TRUE)
par(mfrow=c(nrow=ceiling(dim(coefDF)[1]/2),ncol=2),mar=c(3,3,1,0))
DF_result= data.frame(matrix(NA,nrow=dim(coefDF)[1],ncol=8))
colnames(DF_result)= c('b1','b2','b1_correct','b2_correct','b1_relErr', 'b2_relErr','est_b1','est_b2')
for(i in 1:dim(coefDF)[1]){  
  a1= coefDF$a1[i]; a2= coefDF$a2[i]
  DF_result[i,c(1,2)]= c(a1,a2)
  y_noNoise= a0 + a1*t + a2*ifelse(t>t0,t-t0,0)
  noise= rnorm(length(t),mean=0,sd=0.1*abs(y_noNoise))  #noise relative to data
  y= y_noNoise+noise
  y[y<0]= 0   #make sure no negative values
  plot(t,y,ylim=c(0,1.1*max(y)),main=paste('b1:',a1,', b2:',a2)) #change naming: 'a' are simulated, 'b' estimated params
  
  DF1= twoLine_fit(t,y)
  
  b1_correct= ifelse(mean(DF1$iF1)>0.8 & a1>0 | mean(DF1$iF1)<0.2 & a1==0, 1, 0) #b1 correctly classified
  b1_relErr= abs(mean(DF1$b1)-a1)/abs(a1+1E-3)  #relative error of mean estimate; w. flooring
  b2_correct= ifelse(mean(DF1$iF2)>0.8 & a2<0 | mean(DF1$iF2)<0.2 & a2==0, 1, 0) #b1 correctly classified
  b2_relErr= abs(mean(DF1$b2)-a2)/abs(a2+1E-3)  #relative error of mean estimate
  DF_result[i,3:8]= c(b1_correct,b2_correct,b1_relErr,b2_relErr,median(DF1$b1),median(DF1$b2))
  
  lines(t, mean(DF1$b0) + mean(DF1$b1)*t + mean(DF1$b2) *ifelse(t>mean(DF1$t0),t-mean(DF1$t0),0), col='blue')
}

#manually reformat
DF_result_long= cbind.data.frame(
  coef= c(DF_result$b1,DF_result$b2),
  estCoef= c(DF_result$est_b1,DF_result$est_b2),
  x= factor(c(rep(1,dim(DF_result)[1]),rep(2,dim(DF_result)[1])),levels=c(1,2)),
  y= rep(1:dim(DF_result)[1],times=2),
  correct= factor(c(DF_result$b1_correct,DF_result$b2_correct),levels=c(0,1)),
  relErr= c(DF_result$b1_relErr,DF_result$b2_relErr),
  log10_recipRelErr= log(1/c(DF_result$b1_relErr,DF_result$b2_relErr),10) #log10 of recip. relErr
)

ggplot()+
  geom_point(data=DF_result_long,aes(x=x,y=y,color=correct,size=relErr))+  
  scale_color_manual(limits=as.character(c(0,1)),values=c('red','green'),labels=c('no','yes'))+
  scale_size_continuous(trans='log10')+
  geom_text(data=DF_result_long,aes(x=as.numeric(x)-0.2,y=y, label=coef))+
  theme_minimal()+
  scale_x_discrete(breaks=c(1,2),labels=c('b1','b2'))+
  theme(axis.title= element_blank(),axis.text.y=element_blank())
```

At a noise level of SD = 10%, the parameter estimation did not make any false
positive errors at true rates = 0.